Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-03-30 sources/bmc_template_submission/bmc_article.tex
reviewer 1

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data that
has never been
incorporated into any phylogenetic estimate provides the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases,
with the main goal of building upon published alignments to extend existing
phylogenetic inferences with more data and taxa.
Physcraper updates a starting tree and single locus alignments
with public DNA data, and links the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
Updated and original tree are compared with Robinson-Foulds weighted and unweighted
metrics estimated with Dendropy \cite{sukumaran2010dendropy}, and with a node by node
comparison between the synthetic OpenTree and original and updated tree individually,
using OpenTree's conflict API \cite{redelings2017supertree}.

@@

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment that was used to construct the tree from \cite{gottlieb2005molecular},
using a local GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact that you are discussing running times -->
The Gottlieb2005 tree \cite{gottlieb2005molecular} updated using Physcraper (Figure 2;
Physcraper updated tree from now on) displays all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best ML tree from the RAxML analysis is 99\%
resolved, with 25\% of nodes with bootstrap support $<$ 0.1 and 48\% nodes with
bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
$<$ 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
As comparison with the Physraper updated tree, the Yao2020 tree \cite{yao2020phylogeny}
also contains all 41 distinct taxa
sampled in the Gottlieb2005 tree \cite{gottlieb2005molecular},
while contributing phylogenetic data to 134 additional \textit{Ilex} taxa. From these,
67 taxa are also in the Physcraper updated tree.
While the Yao2020 tree \cite{yao2020phylogeny} was also constructed using
ITS as a marker, their GenBank data is not released yet. Hence, Physcraper was unable
to incorporate 68 taxa that are only on the Yao2020 tree because the DNA data is unavailable.
We also note that Physcraper incorporates 18 \textit{Ilex} taxa that are not in the Yao2020
tree \cite{yao2020phylogeny}. These taxa appear nested among other \textit{Ilex} species
and visual inspection of the DNA sequences suggests they are correctly assigned as \textit{Ilex}.
The ITS alignment that underlies the Yao2020 tree was constructed without any tool to
scrape GenBank \cite{yao2020phylogeny}, which could explain why Physcraper was able to
incorporate these 18 additional \textit{Ilex} taxa in the updated tree (Figure 2).

@@

Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.
The three sequences that were in the original tree but were not included in
the updated tree by Physcraper in the verification test were excluded because
they contain a 100 bp long gap of unidentified nucleotides (Ns) in GenBank that
was manually curated and trimmed in the original alignment.

@@

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}.
However, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral},
BEAST2 \cite{bouckaert2019beast}, or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases,
with the main goal of building upon published alignments to extend existing
phylogenetic inferences with more data and taxa.
Physcraper updates a starting tree and single locus alignments
with public DNA data, and links the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
Updated and original tree are compared with Robinson-Foulds weighted and unweighted
metrics calculated with Dendropy \cite{sukumaran2010dendropy}, and with a node by node
comparison between the synthetic OpenTree and original and updated tree individually,
using OpenTree's conflict API \cite{redelings2017supertree}.

@@

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment that was used to construct the tree from \cite{gottlieb2005molecular},
using a local GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact that you are discussing running times -->
The Gottlieb2005 tree \cite{gottlieb2005molecular} updated using Physcraper (Figure 2;
Physcraper updated tree from now on) displays all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best ML tree from the RAxML analysis is 99\%
resolved, with 25\% of nodes with bootstrap support $<$ 0.1 and 48\% nodes with
bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
$<$ 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
As comparison with the Physcraper updated tree, the Yao2020 tree \cite{yao2020phylogeny}
also contains all 41 distinct taxa
sampled in the Gottlieb2005 tree \cite{gottlieb2005molecular},
while contributing phylogenetic data to 134 additional \textit{Ilex} taxa. From these,
67 taxa are also in the Physcraper updated tree.
While the Yao2020 tree \cite{yao2020phylogeny} was also constructed using
ITS as a marker, their GenBank data is not released yet. Hence, Physcraper was unable
to incorporate 68 taxa that are only on the Yao2020 tree because the DNA data is unavailable.
We also note that Physcraper incorporates 18 \textit{Ilex} taxa that are not in the Yao2020
tree \cite{yao2020phylogeny}. These taxa appear nested among other \textit{Ilex} species
and visual inspection of the DNA sequences suggests they are correctly assigned as \textit{Ilex}.
The ITS alignment that underlies the Yao2020 tree was constructed without any tool to
scrape GenBank \cite{yao2020phylogeny}, which could explain why Physcraper was able to
incorporate these 18 additional \textit{Ilex} taxa in the updated tree (Figure 2).

%% Maybe show conflict results? and then discuss in the discussion.

@@

Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.
The three sequences that were in the original tree but were not included in
the updated tree by Physcraper in the verification test were excluded because
they contain a 100 bp long gap of unidentified nucleotides (Ns) in GenBank that
was manually curated and removed from in the original alignment.

These sequences do appear in the Physcraper results in the list of matches from GenBank which did not fit the sequence length cutoffs set in the configuration. 
This ``seqlen\_mismatch.txt'' file includes the accession number, taxon, and sequence length of all sequences culled based on sequence length.

We encourage users to look at this file, and potentially modify their sequence length cutoff parameters based on the excluded sequences.
The default relative sequence length thresholds are arbitrary, but we hope that by making the process of locating homologous sequences online reproducible, and tracking what cutoffs are used, we make it easier for researchers to delve into the effect that different choices have on their inferences. 
This is in contrast to ‘manual’ searches for taxa, where similarly arbitrary cutoffs are applied, but are difficult to trace.
As many studies have shown \cite{huang_what_2009} the effect of missing data can be enigmatic, and interact with the true phylogenetic relationships for the data set at hand. 
There is not currently strong support in the literature for any cutoffs, and rather than prescribe specific approaches, we encourage users to explore the effects of different choices on inference.

Organellar genome sequences, such as chloroplasts and mitochondria will also generally be excluded from automatic addition based on length cutoffs.
Multiple sequence alignment of loci of drastically different lengths is unfeasible, and we have found in testing often returns incorrect results, splitting shorter sequences with many long gaps to align with exact matches across the entire longer locus. 
While it would be possible to directly extract the blast match from genomes, this match would exclude potentially homologous flanking regions which are not matched by blasts local search algorithm, but may be important for phylogenetic inference.
Instead we list the accession numbers for these matches in the seqlen\_mismatch file, for users to assess appropriate homologous regions to their alignment of interest.

Single gene phylogenies with very high numbers of taxa may lack sufficient signal for accurate resolution \cite{morel_phylogenetic_2020}.
The Physcraper workflow avoids some of these challenges, by focussing on ingroup taxa of an exisiting phylogeny, using markers that have been assessed and proven apporpraite for that phylogenetic scope in past publictions.
In addition, we thin alignments by removing identical sequences, and by setting a maximum number of sequences per taxon.
Nonetheless, it is encumbent on users to assess their final inference with respect to statistical support and biological plausibility.

Most of the default thresholds parameter values and software choices in the Physcraper pipeline are arbitrary. 
We hope that by making the process of locating homologous sequences online reproducible, 
and tracking what cutoffs are used, we make it easier for researchers to delve into the effect that different choices have on their inferences.
In addition, by providing the output files at each step of the analysis, 
it is straightforward to assess how changing parameter and software choices do or do not drive differences in phylogenetic inference.
By gathering the sequences, and making the unaligned files easily available to users, researchers can compare if applying any alternate alignment tool of their choice affects inferences. 
Once sequences are aligned, they can apply and compare inferences from any phylogenetics software.

@@

Physcraper generates gene trees, which individually do not capture the full complexity
of species' evolutionary history \cite{song2012resolving}.
However, Physcraper facilitates
gathering alignments and gene trees for multiple loci from a group of interest,
that together can be used to reconstruct species trees taking into account coalescent
processes with ASTRAL \cite{mirarab2014astral} or SVD Quartets \cite{chifman2014quartet}).
Rigorous analyses of multiple loci allows for more complex evolutionary models
than analyses of large genomic data sets, and can provide better evolutionary estimates.
For example \cite{zhu_complexity_2021} show that when applying coalescent models, there is more information in two genes of 300 bp each than in 600 independent sites 
We have included a script (``multi_locus.py'') and examples of how to automatically merge the outputs of Physcraper runs at different loci into input files for ASTRAL, SVD Quartets, or as concatenated alignments for concatenated analyses.

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
These taxonomic links, and comparisons to existing published phylogenies in the OpenTree data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses, and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences. 
The estimated gene tree itself is an evolutionarily explicit way to visualize gene evolution, which in concert with taxonomic labelling can reveal paralogy.
Our conflict analysis tool informs the users of whether their tree contains major conflicts with established taxonomy. 
This conflict tool also returns information on if these taxonomic conflicts existed in the original input tree. 
These conflicts may be a sign that taxonomy needs to be updated, or may be a sign that non-homologous sequences have been included in the analysis. 
These taxonomic and phylogenetic conflicts flag regions for the tree for the researcher to more closely examine and assess homology.

%% Biological plausibility of example??
